---
title: "baia.seurat.helpers Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{baia.seurat.helpers Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# baia.seurat.helpers Workflow

This vignette explains how the package functions fit together in a BAIA
analysis workflow. The code chunks are marked `eval = FALSE` because the
package does not ship project datasets.

## Workflow Semantics

This package is the generic Seurat helper layer for BAIA single-cell projects and should be used before moving into CITE-seq, DE, pathway, or CellChat-specific analysis.

The package is designed to be used in this progression:

- Sample-sheet helpers
- Seurat object compatibility and merge preparation
- QC annotation and diagnostics
- QC plot constructors
- Visualization utilities

## Sample Metadata And QC Setup

A typical BAIA preprocessing flow starts by reading the sample sheet,
extracting stable sample identifiers, and adding QC metadata that will
be reused by later modules.

```{r, eval=FALSE}
library(baia.seurat.helpers)
samples <- read_samples_sheet("metadata/samples.tsv")
sample_ids <- get_sample_IDs(samples, verbose = FALSE)
so <- compute_percent_mito(so, mitoGenes = grep("^MT-", rownames(so), value = TRUE))
so <- scater_qc(so)
```

## QC Diagnostics

The QC plotting helpers are legacy BAIA utilities. They assume
metadata fields such as `sample`, `nCount_RNA`, `nFeature_RNA`,
`mitoRatio`, or `log10GenesPerUMI` are already present.

```{r, eval=FALSE}
qc_distri_plot(so, "nCount_RNA")
plot_basic_qc(so, title = "Raw QC")
show_qc_plots(so)
```

## Merge Preparation And Visualization

Before merging or handing the object downstream, the helper layer can
drop non-RNA assays, embeddings, graphs, and SCT leftovers so the
object is in a predictable state.

```{r, eval=FALSE}
so <- factor_resolution(so)
merge_ready <- prepare_so_merging(so)
two_groups_dimplot(merge_ready, grp1 = "DHF", grp2 = "DF", grpcol = "WHO_classification")
```

## Full Function Inventory

The following function families summarize the intended purpose of every
exported helper in the package.

### Sample-sheet helpers

| Function | Intended purpose | Minimal example |
| --- | --- | --- |
| `read_samples_sheet()` | Read a BAIA TSV sample sheet whose first header line may start with `#`, returning a plain data frame with character sample IDs preserved. | `read_samples_sheet("metadata/samples.tsv")` |
| `get_sample_IDs()` | Extract sample IDs from a sample sheet data frame or directly from the TSV path. | `get_sample_IDs("metadata/samples.tsv", verbose = FALSE)` |

### Seurat object compatibility and merge preparation

| Function | Intended purpose | Minimal example |
| --- | --- | --- |
| `so_version()` | Return the major Seurat object version as `"4"` or `"5"` and fail early on unsupported objects. | `so_version(so)` |
| `prepare_so_merging()` | Strip a Seurat object down to RNA counts and metadata needed for later merging, removing assays, embeddings, graphs, and SCT leftovers. | `prepare_so_merging(so)` |
| `prepare_so_merging_Seurat4()` | Legacy Seurat v4-specific merge preparation helper with optional verbose logging. | `prepare_so_merging_Seurat4(so, verbose = TRUE)` |
| `factor_resolution()` | Convert clustering resolution columns in `meta.data` into ordered factors for cleaner downstream plotting and table handling. | `factor_resolution(so)` |

### QC annotation and diagnostics

| Function | Intended purpose | Minimal example |
| --- | --- | --- |
| `compute_percent_mito()` | Add `percent.mito` and `percent.mito.cl` metadata based on a supplied vector of mitochondrial gene names. | `compute_percent_mito(so, mitoGenes = grep("^MT-", rownames(so), value = TRUE))` |
| `scater_qc()` | Compute `scater`-style per-cell QC metrics and outlier flags and append them to Seurat metadata. | `scater_qc(so)` |
| `test_cell_cycle_effect()` | Run BAIA's standard cell-cycle scoring and PCA diagnostic workflow to judge whether cell-cycle effects need explicit handling. | `test_cell_cycle_effect(so, "sampleA", s_genes, g2m_genes, showplots = TRUE)` |

### QC plot constructors

| Function | Intended purpose | Minimal example |
| --- | --- | --- |
| `histogram_n_cells()` | Plot the number of cells per sample. | `histogram_n_cells(so)` |
| `density_plot_n_umis()` | Plot the distribution of UMIs per cell, split by sample. | `density_plot_n_umis(so)` |
| `boxplot_n_umis()` | Plot sample-wise boxplots of log10 UMIs per cell. | `boxplot_n_umis(so)` |
| `density_plot_n_genes()` | Plot the distribution of detected genes per cell across samples. | `density_plot_n_genes(so)` |
| `boxplot_n_genes()` | Plot sample-wise boxplots of log10 detected genes per cell. | `boxplot_n_genes(so)` |
| `density_plot_mito_ratio()` | Plot the distribution of mitochondrial ratio across samples. | `density_plot_mito_ratio(so)` |
| `density_plot_complexity()` | Plot transcriptional-complexity distributions such as `log10GenesPerUMI`. | `density_plot_complexity(so)` |
| `dotplot_n_umis_genes_mito()` | Plot UMIs versus genes, colored by mitochondrial ratio, as a compact joint QC view. | `dotplot_n_umis_genes_mito(so)` |
| `qc_distri_plot()` | Build a generic histogram for one named QC metric present in `meta.data`. | `qc_distri_plot(so, "nFeature_RNA")` |
| `plot_basic_qc()` | Draw the BAIA basic QC three-panel violin and jitter view for mitochondrial percentage, genes, and UMIs. | `plot_basic_qc(so, title = "Raw QC")` |
| `plot_highest_expressed()` | Show the highest-expressed genes as horizontal boxplots to identify dominating features. | `plot_highest_expressed(so, nfeat2plot = 20)` |
| `show_qc_plots()` | Print the standard sequence of legacy BAIA QC plots in one call. | `show_qc_plots(so)` |

### Visualization utilities

| Function | Intended purpose | Minimal example |
| --- | --- | --- |
| `two_groups_dimplot()` | Highlight two metadata-defined groups on the same Seurat dimensionality reduction plot. | `two_groups_dimplot(so, grp1 = "DHF", grp2 = "DF", grpcol = "WHO_classification")` |
| `split_print_legend()` | Print a ggplot without its legend and then draw the legend separately, useful for manual report layout. | `split_print_legend(plt)` |

## Relationship To Other BAIA Packages

- `baia.citeseq` builds on this package for generic QC helpers such as
  `compute_percent_mito()`.
- `baia.pseudobulk.deseq` assumes the Seurat objects have already been prepared
  and QCed upstream.
- `baia.ext.cellprop.plots` and `baia.cellchat` consume objects that typically
  originate from this preprocessing layer.


## See Also

- `?baia.seurat.helpers` for the package-level help page
- the README for repository-oriented installation and versioning context
